---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(lmtest)
library(car)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
hist(lab6_data$ndvi_20_med)
hist(lab6_data$maxtemp_20_med)
hist(lab6_data$mintemp_20_med)
hist(lab6_data$rain_20_sum)
hist(lab6_data$pop_20)
```

None of these variables are normally distributed. NDVI and rainfall are moderately right-skewed, maxtemp and mintemp are moderately left-skewed, and population is extremely heavily right skewed.

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
tm_shape(lab6_data)+
  tm_polygons("ndvi_20_med", style = "jenks")
tm_shape(lab6_data)+
  tm_polygons("maxtemp_20_med", style = "jenks")
tm_shape(lab6_data)+
  tm_polygons("mintemp_20_med", style = "jenks")
tm_shape(lab6_data)+
  tm_polygons("rain_20_sum", style = "jenks")
tm_shape(lab6_data)+
  tm_polygons("pop_20", style = "jenks")
```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

NDVI values are highest along the southeast coast and Tasmania and still very high along the northwest, northern, and southwestern coasts NDVI values very low across most of Australia, corresponding to the massive Outback which takes up most of its landmass. Rainfall seems to be strongly correlated to the NDVI values, with the southwestern coast seeming to be an outlier in having low rainfall but high NDVI values. And while most areas with high NDVI values have low population, all areas of high population are found in areas with high NDVI values, while no areas with low NDVI values have a high population. Maxtemp and mintemp are only slightly correlated with NDVI. The primary factor at play there appears to be that northern areas (those closest to the equator) are warmer, with NDVI being correlated with slightly moderated temperatures when comparing across areas at the same latitude.

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
dependent_var<-"ndvi_20_med"
independent_vars<-c("maxtemp_20_med", "mintemp_20_med", "rain_20_sum", "pop_20")

models<-list()
summaries<-list()

for(var in independent_vars){
  formula<-as.formula(paste(dependent_var, "~", var))
  
  model<-lm(formula, data = lab6_data)
  models[[var]]<-model
  summaries[[var]]<-summary(model)
  
  cat("\nModel for", var, ":\n")
  print(summaries[[var]])
}
```

The directions and magnitude for each univariate model were shown in the coefficients, better described as the slope of the linear equation. The coefficients for the min and max temperature correlations were negative (-0.012 and -0.013 respectively), while rainfall and population had much smaller positive coefficients (9.124 x 10^-7 and 1.500 x 10^6 respectively). Because NDVI values only range from -1 to 1, these correlations have a larger magnitude than it appears, but they are still relatively small.

The correlation strength for each uni variate model varied widely. The strongest correlation by far was between rainfall and NDVI (Rsq = 0.43), with the weakest being between population and NDVI (Rsq = 0.12). Interestingly, max temp and NDVI had a stronger correlation than min temp and NDVI (Rsq = 0.21 and Rsq = 0.14, respectively). Additionally, all coefficients across all correlations were shown as highly significant as shown by the significance codes.

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
dependent_var1<-"ndvi_20_med"
independent_vars1<-c("maxtemp_20_med", "rain_20_sum", "pop_20")

formula1<-as.formula(paste(dependent_var1, "~", paste(independent_vars1, collapse = " + ")))
multivar_model<-lm(formula1, data = lab6_data)
summary(multivar_model)
```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

The multivariate model shows a positive coefficient for all variables except min temp. However, magnitudes for annual rainfall and population are quite small, with an order of magnitude at or below 10^-6. The magnitude agnitude of max temp is larger at 10^-3 . The F-statistic value of 312.1 and the p-value of less than 2.2 x 10^-16 shows that the model is useful, and the R-squared value of 0.64 tells us that about 64% of the variance in the data is explained by our model.

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
residuals<-residuals(multivar_model)
hist(residuals)

lab6_data<-lab6_data %>%
  mutate(ndvi_20_med_pred = predict(multivar_model)) %>%
  mutate(ndvi_20_pred_diff = ndvi_20_med_pred - ndvi_20_med)

tm_shape(lab6_data)+
  tm_polygons("ndvi_20_pred_diff", style = "jenks")
```

The residuals appear strongly left-skewed. The model seems to underpredict NDVI values along the eastern coast of Australia and in Tasmania, overpredict in clusters of southern and northwestern Australia, and accurately predict in most of the country.

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
vif(multivar_model)
bptest(multivar_model)
```

The vif test for multicollinearity shows that total rainfall, population, and max temp should be included in our model. This is because these values were below 5, which is the cutoff for significance. The Breusch-Pagan test shows a p-value of 3.64 x 10^-14, which is well below the p = 0.05 value required to overturn the null hypothesis that the data are not heteroskedastic. This means we can conclude that the data do not follow constant variance, making our model unreliable.

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

Our model explains about 66% of the variance in the data, with reasonably good predicitions for much of Australia, and all our variables seem to have relevance to our prediction. However, underprediction and overprediction seem to be concentrated in specific areas, begging the question of whether more or different variables are needed to create an accurate model of NDVI values for Australia. Based on this information, I think the results of this model are meaningful but in need of improvement via the addition of more variables.

**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

I received help from a friend on questions 7 and 8 concerning concepts, and help from ChatGPT on questions 4 and 5 with debugging.

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

I do not feel confident about my work on this lab, as I was confused by a number of key concepts such as heteroskedacisity, and Breusch-Pagan tests.

**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}

```




#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

